---
title: "Avoidance-Attraction Ratios in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Avoidance-Attraction Ratios in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Avoidance-Attraction Ratios were first developed by Parsons et al. 2016 to investigate the effects of humans and their favorite companion, dogs, on wildlife. Since then Avoidance-Attraction Ratios have grown in popularity due to their ability to gain insights into species interactions at a fine temporal scale.

This guide will implement the functions in the AvoidAttract package and perform an analysis of species interactions using Avoidance-Attraction Ratios having bobcat (*Lynx rufus*), coyotes (*Canis latrans*), and swift fox (*Vulpes velox*) as the predator species. The data that we will use is integrated in the package and is from a camera trapping project in western Kansas, source: Palomo-Munoz et al. 2023.

Lets get started.

### Package Install


```{r setup}
# Install the package
# If needed install.package("devtools")
devtools::install_github("dpearcetamu/AvoidAttract", quiet = TRUE)

# Load the package
library(AvoidAttract)
```

Now that the package is installed and loaded, let's take a look at the data.

### Exploring the Data

```{r data documentation}
# If you want to take a closer look use this code to read the dataset documentation, 
# ?KScams_dat
```

Okay, now that we've seen the details of the camera trapping data, lets load it into the environment and take a closer look.

```{r reading in data}
# Reading in the data
data("KScams_dat")

# Let's take a look
head(KScams_dat)
```

Great! We can see that the data has detection's of species at certain sites with a specific date and time for that observation. Let's see how many different species were found.

```{r looking into the data}
# Let's see what the column names are 
names(KScams_dat)

# What were the unique species using the "Common_name" column
unique(KScams_dat$Common_name)
```

Looking to see how many observations of each species there are in the data set using the spp_sum function from the package.

```{r species observations}
# spp_sum requires the data input and a name column and provides a frequency table of observations
spp_sum(KScams_dat, name_col = "Common_name")

```

That's a lot of observations! Let's take a look at a few of the species that had a lot of observations and their interactions with coyotes using the AAR function.

### Avoidance-Attraction Ratios

```{r AAR function}
# First we have to format the DateTime column from a character string to a date time 
# This is an important step since the function won't know how to read the date and time
# Make sure to modify the format if you are using a different dataset!!!
KScams_dat$DateTime  <- as.POSIXct(KScams_dat$DateTime ,  
                                   tryFormats = "%m/%d/%Y %H:%M:%OS")

# Let's look at some of these species!

# ----------------------------------------
#
#       Bobcats as predators
#
# ----------------------------------------

# Starting with the Black-tailed jackrabbit 
jrabbit_bobcat <- AAR(data = KScams_dat, 
                   speciesA = "Black-tailed jackrabbit", speciesB = "Bobcat",
                   species_col = "Common_name", datetime_col = "DateTime", 
                   site_col ="Site", unitTime = "hours")

# The AAR function outputs several results. Let's take a look at just the total summary.
print(jrabbit_bobcat$total_summary)

# Okay we can see that this outputs the mean values 
# for all interaction events and the interaction ratios

# The AAR function also returns a list of how many times those interactions occurred.
print(jrabbit_bobcat$event_count)

# That's quite a few counts for each interaction.
# Lets save just the ratios to use later on.

# Saving only the T2/T1 and T4/T3 ratios 
jrabbit_bobcat_ratios <- jrabbit_bobcat$total_summary[c(5,6)]

# Now the other species

# Black-tailed Prarie Dog
pdog_bobcat <- AAR(data = KScams_dat, 
                speciesA = "Black-tailed Prarie Dog", speciesB = "Bobcat",
                species_col = "Common_name", datetime_col = "DateTime", 
                site_col ="Site", unitTime = "hours")

# Okay there weren't enough time events to calculate ratios.
# Let's take a closer look.
print(pdog_bobcat$event_count)

# There were no T2 or T4 events and only one T1 event for the prarie dog 
# and bobcat interaction. Since we are trying to compare differences between
# predator species we won't look into prairie dog's interactions with 
# coyote and swift fox.

# Eastern cottontail rabbit
crabbit_bobcat <- AAR(data = KScams_dat, 
                  speciesA = "Eastern cottontail rabbit", speciesB = "Bobcat",
                  species_col = "Common_name", datetime_col = "DateTime", 
                  site_col ="Site", unitTime = "hours")

crabbit_bobcat_ratios <- crabbit_bobcat$total_summary[c(5,6)]


# White-tailed Deer
wtd_bobcat <- AAR(data = KScams_dat, 
               speciesA = "White-Tailed Deer", speciesB = "Bobcat",
               species_col = "Common_name", datetime_col = "DateTime", 
               site_col ="Site", unitTime = "hours")


wtd_bobcat_ratios <- wtd_bobcat$total_summary[c(5,6)]


# ----------------------------------------
#
#       Coyote as predators
#
# ----------------------------------------

# Starting with the Black-tailed jackrabbit 
jrabbit_coyote <- AAR(data = KScams_dat, 
                   speciesA = "Black-tailed jackrabbit", speciesB = "Coyote",
                   species_col = "Common_name", datetime_col = "DateTime", 
                   site_col ="Site", unitTime = "hours")

# Saving only the T2/T1 and T4/T3 ratios 
jrabbit_coyote_ratios <- jrabbit_coyote$total_summary[c(5,6)]

# Eastern cottontail rabbit
crabbit_coyote  <- AAR(data = KScams_dat, 
                  speciesA = "Eastern cottontail rabbit", speciesB = "Coyote",
                  species_col = "Common_name", datetime_col = "DateTime", 
                  site_col ="Site", unitTime = "hours")

crabbit_coyote_ratios <- crabbit_coyote$total_summary[c(5,6)]


# White-tailed Deer
wtd_coyote <- AAR(data = KScams_dat, 
               speciesA = "White-Tailed Deer", speciesB = "Coyote",
               species_col = "Common_name", datetime_col = "DateTime", 
               site_col ="Site", unitTime = "hours")


wtd_coyote_ratios <- wtd_coyote$total_summary[c(5,6)]


# ----------------------------------------
#
#       Swift fox as predators
#
# ----------------------------------------

# Starting with the Black-tailed jackrabbit 
jrabbit_fox <- AAR(data = KScams_dat, 
                   speciesA = "Black-tailed jackrabbit", speciesB = "Swift Fox",
                   species_col = "Common_name", datetime_col = "DateTime", 
                   site_col ="Site", unitTime = "hours")

# Saving only the T2/T1 and T4/T3 ratios 
jrabbit_fox_ratios <- jrabbit_fox$total_summary[c(5,6)]

# Eastern cottontail rabbit
crabbit_fox  <- AAR(data = KScams_dat, 
                  speciesA = "Eastern cottontail rabbit", speciesB = "Swift Fox",
                  species_col = "Common_name", datetime_col = "DateTime", 
                  site_col ="Site", unitTime = "hours")

crabbit_fox_ratios <- crabbit_fox$total_summary[c(5,6)]


# White-tailed Deer
wtd_fox <- AAR(data = KScams_dat, 
               speciesA = "White-Tailed Deer", speciesB = "Swift Fox",
               species_col = "Common_name", datetime_col = "DateTime", 
               site_col ="Site", unitTime = "hours")


wtd_fox_ratios <- wtd_fox$total_summary[c(5,6)]


```

### Data Wrangling

Okay, we need to organize the ratios for each prey species into one dataframe for analysis.

```{r combine summaries}
# Load libraries
# If needed install.packages("tidyverse")
options(tidyverse.quiet = TRUE)
library(tidyverse)

# Combining jack rabbit ratio data 
jrabbit_ratios <- t(data.frame(
                    # T2/T1 ratios
                    bobcat_T2_T1 = jrabbit_bobcat_ratios["T2/T1"], 
                    coyote_T2_T1 = jrabbit_coyote_ratios["T2/T1"],
                    fox_T2_T1 = jrabbit_fox_ratios["T2/T1"],
                    
                    # T4/T3 ratios
                    bobcat_T4_T3 = jrabbit_bobcat_ratios["T4/T3"],
                    coyote_T4_T3 = jrabbit_coyote_ratios["T4/T3"],
                    fox_T4_T3 = jrabbit_fox_ratios["T4/T3"]))


# Convert the matrix to a data frame
jrabbit_ratios <- as.data.frame(jrabbit_ratios)

# Change row names to a new column
jrabbit_ratios <- tibble::rownames_to_column(jrabbit_ratios, var = "RowName")

# We have to rename the columns to match the values correctly
colnames(jrabbit_ratios) <- c("Interaction", "Value")

# Adding a ratio column
jrabbit_ratios$Ratio <- c("T2/T1", "T2/T1", "T2/T1",
                          "T4/T3", "T4/T3", "T4/T3")

# Checking to make sure the Ratios are named correctly with current interaction
# names
print(jrabbit_ratios)

# Naming interactions by predator
jrabbit_ratios$Interaction <- c("Bobcat", "Coyote","Swift Fox",
                                "Bobcat", "Coyote", "Swift Fox")

# Checking to see if it looks correct
print(jrabbit_ratios)

# Combining cottontail rabbit             
crabbit_ratios <- t(data.frame(
                  # T2/T1 ratios
                  bobcate_T2_T1 = crabbit_bobcat_ratios["T2/T1"], 
                  coyote_T2_T1 = crabbit_coyote_ratios["T2/T1"],
                  fox_T2_T1 = crabbit_fox_ratios["T2/T1"],
                  
                  # Column for T4/T3 ratios
                  bobcate_T4_T3 = crabbit_bobcat_ratios["T4/T3"],
                  coyote_T4_T3 = crabbit_coyote_ratios["T4/T3"],
                  fox_T4_T3 = crabbit_fox_ratios["T4/T3"])) 


# Convert the matrix to a data frame
crabbit_ratios <- as.data.frame(crabbit_ratios)

# Change row names to a new column
crabbit_ratios <- tibble::rownames_to_column(crabbit_ratios, var = "RowName")

# We have to rename the columns to match the values correctly
colnames(crabbit_ratios) <- c("Interaction", "Value")

# Adding a ratio column
crabbit_ratios$Ratio <- c("T2/T1", "T2/T1", "T2/T1",
                         "T4/T3", "T4/T3", "T4/T3")
# Naming interactions by predator
crabbit_ratios$Interaction <- c("Bobcat", "Coyote","Swift Fox",
                                "Bobcat", "Coyote", "Swift Fox")


# Combining cottontail rabbit             
wtd_ratios <- t(data.frame(
                # T2/T1 ratios
                bobcate_T2_T1 = wtd_bobcat_ratios ["T2/T1"], 
                coyote_T2_T1 = wtd_coyote_ratios["T2/T1"],
                fox_T2_T1 = wtd_fox_ratios["T2/T1"],
                
                # Column for T4/T3 ratios
                bobcate_T4_T3 = wtd_bobcat_ratios ["T4/T3"],
                coyote_T4_T3 = wtd_coyote_ratios["T4/T3"],
                fox_T4_T3 = wtd_fox_ratios["T4/T3"]))

# Convert the matrix to a data frame
wtd_ratios <- as.data.frame(wtd_ratios)

# Change row names to a new column
wtd_ratios <- tibble::rownames_to_column(wtd_ratios, var = "RowName")

# We have to rename the columns to match the values correctly
colnames(wtd_ratios) <- c("Interaction", "Value")

# Adding a ratio column
wtd_ratios$Ratio <- c("T2/T1", "T2/T1", "T2/T1",
                      "T4/T3", "T4/T3", "T4/T3")

# Naming interactions by predator
wtd_ratios$Interaction <- c("Bobcat", "Coyote","Swift Fox",
                            "Bobcat", "Coyote", "Swift Fox")


```

### Statistical Analysis

Let's followthe analysis laid out by Parsons (2016) using a Wilcoxon test for each of our prey species and their interactions with the predator species and our prey species.

```{r analysis}
# Jack rabbit
# Separate the data into two data frames for T2/T1 and T4/T3
jrabbit_t2t1 <- subset(jrabbit_ratios, Ratio == "T2/T1")
jrabbit_t4t3 <- subset(jrabbit_ratios, Ratio == "T4/T3")

# Run Wilcoxon pairwise tests for T2/T1
jrabbit_wilcox_t2t1 <- pairwise.wilcox.test(jrabbit_t2t1$Value, 
                                            jrabbit_t2t1$Interaction, 
                                            p.adjust.method = "bonferroni")

# Run Wilcoxon pairwise tests for T4/T3
jrabbit_wilcox_t4t3 <- pairwise.wilcox.test(jrabbit_t4t3$Value, 
                                            jrabbit_t4t3$Interaction, 
                                            p.adjust.method = "bonferroni")

# Print the results
print("Jack rabbit Wilcoxon Pairwise Test Results for T2/T1:")
print(jrabbit_wilcox_t2t1)

print("Jack rabbit Wilcoxon Pairwise Test Results for T4/T3:")
print(jrabbit_wilcox_t4t3)

# Cottontail rabbit

# Separate the data into two data frames for T2/T1 and T4/T3
crabbit_t2t1 <- subset(crabbit_ratios, Ratio == "T2/T1")
crabbit_t4t3 <- subset(crabbit_ratios, Ratio == "T4/T3")

# Run Wilcoxon pairwise tests for T2/T1
crabbit_wilcox_t2t1 <- pairwise.wilcox.test(crabbit_t2t1$Value,
                                    crabbit_t2t1$Interaction, 
                                    p.adjust.method = "bonferroni")

# Run Wilcoxon pairwise tests for T4/T3
crabbit_wilcox_t4t3 <- pairwise.wilcox.test(crabbit_t4t3$Value, 
                                    crabbit_t4t3$Interaction, 
                                    p.adjust.method = "bonferroni")

# Print the results
print("Cottontail rabbit Wilcoxon Pairwise Test Results for T2/T1:")
print(crabbit_wilcox_t2t1)

print("Cottontail rabbit Wilcoxon Pairwise Test Results for T4/T3:")
print(crabbit_wilcox_t4t3)

# White-tailed Deer

# Separate the data into two data frames for T2/T1 and T4/T3
wtd_t2t1 <- subset(wtd_ratios, Ratio == "T2/T1")
wtd_t4t3 <- subset(wtd_ratios, Ratio == "T4/T3")

# Run Wilcoxon pairwise tests for T2/T1
wilcox_t2t1 <- pairwise.wilcox.test(wtd_t2t1$Value,
                                    wtd_t2t1$Interaction, 
                                    p.adjust.method = "bonferroni")

# Run Wilcoxon pairwise tests for T4/T3
wilcox_t4t3 <- pairwise.wilcox.test(wtd_t4t3$Value, 
                                    wtd_t4t3$Interaction, 
                                    p.adjust.method = "bonferroni")

# Print the results
print("White-tailed Deer Wilcoxon Pairwise Test Results for T2/T1:")
print(wilcox_t2t1)

print("White-tailed Deer Wilcoxon Pairwise Test Results for T4/T3:")
print(wilcox_t4t3)

```

The results of the Wilcoxon test show that none of the ratios were statistically significant. But that doesn't mean that there isn't a response in attraction or avoidance, only that it wasn't statistically significant. We know that values >1 suggest nonrandom movement between the two species which suggests that there is an interaction. We also know that attraction of species B to species A would result in high T2/T1 ratios, but lower T4/T3 ratios. Avoidance of species A from species B would result in higher ratios of T4/T3.


### Plotting

Let's plot both of the ratios and take a look.



```{r fig.width=8, fig.height=8}
# Let's plot the ratios with ggplot2
# If needed install.packages("ggplot2")
library(ggplot2, quiet = TRUE)

# Plotting Jack rabbit interactions
ggplot(jrabbit_ratios, aes(x = Interaction, y = Value, fill = Ratio)) +
                      geom_col(position = position_dodge(0.8), width = 0.7) +
                      geom_text(aes(label = round(Value, 2)), 
                                position = position_dodge(0.8), vjust = -0.5, size = 4) +
                      scale_fill_manual(values = c("black", "grey"), name = "Ratio") +  
                      labs(x = "Species", y = "Avoidance-Attraction Ratios", 
                           title = "Jack Rabbit Avoidance and Predator Attraction") +
                      theme(axis.text.x = element_text(angle = 45, hjust = 1),
                            panel.background = element_rect(fill = "white"),
                          plot.title = element_text(size = 16)) 

# Plotting Cottontail rabbit interactions
ggplot(crabbit_ratios, aes(x = Interaction, y = Value, fill = Ratio)) +
                      geom_col(position = position_dodge(0.8), width = 0.7) +
                      geom_text(aes(label = round(Value, 2)), 
                                position = position_dodge(0.8), vjust = -0.5, size = 4) +
                      scale_fill_manual(values = c("black", "grey"), name = "Ratio") +  
                      labs(x = "Species", y = "Avoidance-Attraction Ratios", 
                           title = "Cottontail Rabbit Avoidance and Predator Attraction") +
                      theme(axis.text.x = element_text(angle = 45, hjust = 1),
                            panel.background = element_rect(fill = "white"),
                          plot.title = element_text(size = 16))



# Plotting White-tailed Deer interactions
ggplot(wtd_ratios, aes(x = Interaction, y = Value, fill = Ratio)) +
                    geom_col(position = position_dodge(0.8), width = 0.7) +
                    geom_text(aes(label = round(Value, 2)), 
                              position = position_dodge(0.8), vjust = -0.5, size = 4) +
                    scale_fill_manual(values = c("black", "grey"), name = "Ratio") +  
                    labs(x = "Species", y = "Avoidance-Attraction Ratios", 
                         title = "White-tailed Deer Avoidance and Predator Attraction") +
                    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                          panel.background = element_rect(fill = "white"),
                          plot.title = element_text(size = 16)) 

```

We can see from the plots that none of the T2/T1 values are >1. This suggests that none of the predator species are actively seeking out the prey species. But all of the T4/T3 values were >1 which suggests that the prey are avoiding the predator. 


### References:

Parsons, A. W., C. Bland, T. Forrester, M. C. Baker-Whatton, S. G. Schuttler, W. J. McShea, R. Costello, and R. Kays. 2016. The ecological impact of humans and dogs on wildlife in protected areas in eastern North America. Biological Conservation 203:75--88. URL: <https://doi.org/10.1016/j.biocon.2016.09.001>

Palomo‐Munoz, G., T. J. Werdel, C. W. Piper, M. S. Peek, A. M. Ricketts, and A. A. Ahlers. 2023. Spatiotemporal distributions of mammals occurring in an agro‐prairie ecosystem. Ecology 104:e3945 URL: https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.3945

